---
title: "Principal Component Analysis with Interactive Fixed Effect Model"
author: "Erik Loualiche"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
#output:
#  md_document:
#    variant: markdown_github
always_allow_html: yes
vignette: >
  %\VignetteIndexEntry{IV and Interactive fixed effect models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

First load the package

```{r, eval=TRUE, message=F, warnings=F, highlight=TRUE, cache=T, results='hide'}
library(magrittr); library(dplyr);
library(data.table)
library(FixedEffectjlr)
JULIA_HOME <- "/Applications/Julia-0.6.app/Contents/Resources/julia/bin/"
FixedEffect_setup(JULIA_HOME)
```
 
## How to recover principal component analysis from Interacted Fixed Effects

We are going to use the standard dataset:
```{r, eval=T, message = FALSE, warnings = FALSE, highlight=TRUE}
dt <- Ecdat::Cigar %>% data.table
```

### Interactive fixed effects as PC
IFE is similar to a principal component analysis whenever there is no right hand side parameter. The standard specification is:
```{r, eval=TRUE, message=F, warnings=F, highlight=TRUE}
ife <- 
  FixedEffectInteract(dt, lhs = "sales", rhs = "0", 
                      ife = "state+year", rank=1, 
                      fe = "0")
```

To compare with the principal component procedure, we look at loadings and factors first and then the residuals.
```{r, eval=TRUE, message=F, warnings=F, highlight=TRUE}
ife_loadings <- unique(ife$dt_augment[, .(state, ife_loadings1=loadings1)])
ife_factors  <- unique(ife$dt_augment[, .(year, ife_factors1=factors1)])
```

### Comparing with Principal Component Analysis
Before we can use `prcomp` we need to transform the data from long to wide:
```{r, eval=TRUE, message=F, warnings=F, highlight=TRUE}
dt_wide <- dt[, .(state, year, sales) ] %>%
    dcast(year ~ state, value.var = "sales")
```

Now we are able to directly run the PC analysis. Note that IFE does not rescale the variable, so we are going to run it with both centering and rescaling options turned off:
```{r, eval=TRUE, message=F, warnings=F, highlight=TRUE}
pca_res <- prcomp(dt_wide[, -"year"], center=FALSE, scale=FALSE)
# keep loadings and factors separately
pca_loadings <- cbind(unique(dt[, .(state)]), 
                      data.table(pca_res$rotation)[, .(pc_loadings1=PC1, pc_loadings2=PC2) ] )
pca_factors  <- cbind(unique(dt[, .(year)]),  
                      data.table(pca_res$x)[, .(pc_factors1 = PC1, pc_factors2 = PC2) ] )
# remerge to estimate residuals
dt_pca <- dt[, .(state, year, sales) ] %>%
  merge(., pca_loadings, by = c("state") , all.x = T) %>%
  merge(., pca_factors, by = c("year"), all.x = T)
```

Finally we compare loadings, factors and $R^2$:
```{r, eval=TRUE, message=F, warnings=F, highlight=TRUE}
# loadings
dt_loadings <- merge(pca_loadings, ife_loadings, by = c("state"))
# are they proportional to each other (difference in scale)
dt_loadings[, .(state, prop_loadings1 = ife_loadings1/pc_loadings1)] %>% as_tibble
```

We do the same thing for factors
```{r, eval=TRUE, message=F, warnings=F, highlight=TRUE}
# factors
dt_factors <- merge(pca_factors, ife_factors, by = c("year"))
# are they proportional to each other (difference in scale)
dt_factors[, .(year, prop_factors1 = ife_factors1/pc_factors1)] %>% as_tibble
```

Last note that the proportionality factors are actually the inverse of each other:
```{r, eval=TRUE, message=F, warnings=F, highlight=TRUE}
dt_loadings[, .(mean(ife_loadings1/pc_loadings1)) ] * 
  dt_factors[, .(mean(ife_factors1/pc_factors1))]
```


And finally we look to see if residuals are the same:
```{r, eval=TRUE, message=F, warnings=F, highlight=TRUE}
# residuals for only one factor
dt_residuals <- 
  merge(ife$dt_augment[, .(state, year, ife_residuals=residuals)],
        dt_pca[, .(state, year, pc_residuals=sales-pc_loadings1*pc_factors1)],
        by = c("year", "state"))
dt_residuals[, .(ife_residuals = sum(ife_residuals^2), 
                 pc_residuals = sum(pc_residuals^2)) ]
```


## How to do we get the Principal Component Analysis when the factors are demeaned

We start with the PC analysis. Except that this time we allow for a centering:
```{r, eval=TRUE, message=F, warnings=F, highlight=TRUE}
pca_res <- prcomp(dt_wide[, -"year"], center=TRUE, scale=FALSE)
# keep loadings and factors separately
pca_loadings <- cbind(unique(dt[, .(state)]), 
                      data.table(pca_res$rotation)[, .(pc_loadings1=PC1, pc_loadings2=PC2) ] )
pca_factors  <- cbind(unique(dt[, .(year)]),  
                      data.table(pca_res$x)[, .(pc_factors1 = PC1, pc_factors2 = PC2) ] )
# remerge to estimate residuals
dt_pca <- dt[, .(state, year, sales) ] %>%
  merge(., pca_loadings, by = c("state") , all.x = T) %>%
  merge(., pca_factors, by = c("year"), all.x = T)
```


This is similar to removing the within group average, which corresponds to group fixed effects. We obtain the centering in IFE through:
```{r, eval=TRUE, message=F, warnings=F, highlight=TRUE}
ife <- 
  FixedEffectInteract(dt, lhs = "sales", rhs = "0", 
                      ife = "state+year", rank=1, 
                      fe = "state")
ife_loadings <- unique(ife$dt_augment[, .(state, ife_loadings1=loadings1)])
ife_factors  <- unique(ife$dt_augment[, .(year, ife_factors1=factors1)])
```


As in the previous example, we merge loadings, factors and compare $R^2$:
```{r, eval=TRUE, message=F, warnings=F, highlight=TRUE}
# loadings
dt_loadings <- merge(pca_loadings, ife_loadings, by = c("state"))
# are they proportional to each other (difference in scale)
dt_loadings[, .(state, prop_loadings1 = ife_loadings1/pc_loadings1)] %>% as_tibble
# factors
dt_factors <- merge(pca_factors, ife_factors, by = c("year"))
# are they proportional to each other (difference in scale)
dt_factors[, .(year, prop_factors1 = ife_factors1/pc_factors1)] %>% as_tibble
```

We confirm that the proportionality factors are actually the inverse of each other:
```{r, eval=TRUE, message=F, warnings=F, highlight=TRUE}
dt_loadings[, .(mean(ife_loadings1/pc_loadings1)) ] * 
  dt_factors[, .(mean(ife_factors1/pc_factors1))]
```


And finally we look to see if residuals are the same:
```{r, eval=TRUE, message=F, warnings=F, highlight=TRUE}
# residuals for only one factor
dt_residuals <- 
  merge(ife$dt_augment[, .(state, year, ife_residuals=residuals)],
        dt_pca[, .(year, pc_residuals=sales-mean(sales)-pc_loadings1*pc_factors1),
               by = .(state) ],
        by = c("year", "state"))
dt_residuals[, .(ife_residuals = sum(ife_residuals^2), 
                 pc_residuals = sum(pc_residuals^2)) ]
```
